### 2007-01-15 Documentation/part2.tex
JK - Minor corrections in chapter 10

Here the general concept of the AEDL is described and illustrated by an example. In addition, a short introduction to the syntax of the AEDL is given.

@@

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method} (see, \eg, \Ref~\cite{Ohlsson:1999um})  to compute the neutrino oscillation probabilities and divide the matter density profile into layers of constant matter density. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained by the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even if only a small number matter density steps is used. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

While it is comparatively simple to define a general neutrino source 
and to compute the oscillation physics, the general properties of a detector simulation are much more complicated. The basic assumption in building an abstract detector description is \emph{linearity}, \ie , that two neutrino events do not interfere with each other. Furthermore it is assumed that all information on the oscillation physics 
is given by the \emph{reconstructed} flavor and energy of a 
neutrino event. The term ``reconstructed'' implies that the well-defined energy of the incident neutrino, which can not be directly observed, translates via secondary particles and the detection properties into a distribution of possible energy values. This process is illustrated in \figu{distro} for the energy variable. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.6\textwidth]{mapping}
\end{center}
\mycaption{\label{fig:distro} A detector maps a true parameter value onto
a distribution of reconstructed parameter values. This is illustrated here for there energy.}
\end{figure}
% 
The same, in principle, applies to the nature of the neutrino flavor. However, in this case only discrete values are applicable. Note that the reconstructed neutrino energy and the neutrino flavor are the only observables in \GLOBES .

@@

In order to implement a experiment definition including various
sources of systematical errors, we use several abstraction levels. 
The first level is the so-called ``channel'', \index{norm}{Channel}
 which is the link between 
the oscillation physics and the detection properties for a specfific oscillation pattern (\cf, \figu{channel}). A channel specifies the mapping of a specific neutrino flavor produced by the source onto a reconstructed neutrino flavor.
For example, a muon neutrino oscillates into an electron neutrino and subsequently interacts via quasi-elastic charged current scattering. The measured energy and direction of
the secondary electron in the detector then allows to reconstruct the neutrino energy. The connection from the source flux of the muon neutrino, via the  probability to appear as a electron neutrino, to its detection properties (such as cross sections and energy smearing) is encapsulated into the channel.

@@

The channels are the building blocks for the so-called ``rules''.
\index{norm}{Rule} In general, a rule consists of one or more ``signal'' and ``background'' oscillation channels, which are normalized with efficiencies
(\cf, \figu{rule}). The event numbers from these channels are added {\em before} the $\Delta \chi^2$-value is calculated.\footnote{Note that
in this manual, the $\chi^2$ and $\Delta \chi^2$ are equal, since
for simulated data $\Delta \chi^2 = 0$ at the best-fit point. Thus, we
are using $\chi^2$ and $\Delta \chi^2$ as equal quantities.} In addition, each rule implements an {\em independent} systematics, such as signal and background normalization errors. Eventually, each rule gives a $\Delta \chi^2$-value, and the total $\Delta \chi^2$ of one experiment is obtained by adding the $\Delta \chi^2$'s of all rules (\cf, \figu{experiment}). 
 An example for a rule could look like this: We want to detect electron
  neutrino appearance (``signal''), where the overall efficiency for 
  quasi-elastics electron neutrino events is $0.4$. There is a fraction of 
  $0.01$ of all neutral current events which are mis-identified as 
  quasi-elastic electron neutrino events (``background''). The neutral 
  current fraction is only known within $10\%$ (``background uncertainty'') 
  and there is
an energy scale uncertainty of $100\,\mathrm{MeV}$ (``energy calibration error'').
All this systematics is independent of the other rules.  Thus, a rule connects the event rates to the calculation of a $\Delta \chi^2$ which properly includes systematical errors. The resulting $\Delta \chi^2$ is then the starting point for the oscillation physics analysis. Note again that
\begin{itemize}
\item
 Within each rule the event numbers are added.
\item
 Within each rule the systematics is treated independently from the other rules.
\item
 For each rule the $\Delta \chi^2$ is computed; the $\Delta \chi^2$'s from all rules are added.
\end{itemize}

@@

Experiments are in \GLOBES\ defined by the Abstract Experiment Definition Language (AEDL). The experiment definition is written into a text file using the AEDL syntax. Currently, a number of pre-defined experiment definition files are provided with \GLOBES , which have to be modified manually in order to define new experiments.  The application software then uses this text file to initialize the experiment, where other secondary files might read for source fluxes, cross sections \etc . In this section, we show the definition of a very simple neutrino factory in AEDL, where we do not go into details. In the next chapter, we will discuss each of the individual steps in detail.

@@

originating from stored $\mu^+$'s. This achieved by setting the {\tt @builtin} variable to $1$. Next, we specify the muon energy to be $50\,\mathrm{GeV}$ by the {\tt @parent\_energy} variable. We assume 

@@

Then we specify the file which contains the cross sections we want to 

@@

one element, because we only use one density layer: We initialize a baseline length of $3000 \, \mathrm{km}$ with a constant matter density of $3.5 \, \mathrm{g/cm^3}$. 

@@

above defined channel named {\tt \#appearance} with an constant overall

@@

The background error variable is defined such as the signal error variable, \ie , we have a $20\%$ background uncertainty and a very small energy calibration error. In addition, the systematics treatment is specified

@@

The here defined experiment represents a first simplified version of a neutrino factory experiment. It still lacks the correct energy dependence of the efficiencies, the antineutrino disappearance channel, and the channels and rules for the symmetric operation with $\mu^-$ stored. However, it may serve as a simple, introductory example. In the next chapter, we will demonstrate that the AEDL is much more powerful than illustrated here.

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages.  In addition, the version of the minimum
\GLOBES\ version the \AEDL\ file is running with has to be defined by {\tt \$version}, such as
{\tt \$version="3.0.0"}.
\index{aedl}{version@{\tt \$version}}
%
Comments can be used such as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */ \\
// Another comment
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$.\footnote{The upper limit is 
only there for safety reasons, the memory is allocated dynamically.} If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environments in AEDL:
\begin{quote}
{\tt nuflux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a declaration without definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). In order to facilitate the the mapping from names in AEDL to indices
in C there are two functions \GLB{NameToValue} and \GLB{ValueToName} which
make this transition (see \Sec~\ref{sec:aedl_names}, 
page~\pageref{sec:aedl_names}).

@@

\index{aedl}{include@{\tt include}}
As another feature of AEDL one can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "Exp1.glb"\\
        \#NEXT\#\\
        include   "Exp2.glb"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt Exp1.glb} and {\tt Exp2.glb}. Note that one has 
to use quotation marks for filenames in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively by two separate calls of 
\GLB{InitExperiment}.

Furthermore, one can define constants such as
\begin{quote}
{\tt
Pi = 3.141
}
\end{quote}
These constants can not only be defined within one \AEDL\ file, but also
by the calling C program, which allows to use a simple but powerful variable
substitution mechanism as described in~\Sec~\ref{sec:aedlparams}.
\index{norm}{AEDL@\AEDL!external parameters}

In addition, some simple algebraic manipulations are possible, such as
\begin{quote}
{\tt
Pi+1\\
\verb+Pi^2+\\
sin(Pi/2)
}
\end{quote}
The following mathematical functions from {\tt <math.h>} are available: 
{\tt sin}, {\tt cos}, {\tt tan}, {\tt asin}, {\tt acos}, {\tt atan}, 
{\tt log}, {\tt log10}, {\tt exp}, {\tt sqrt}.
\index{aedl}{sin@{\tt sin}}
\index{aedl}{cos@{\tt cos}}
\index{aedl}{tan@{\tt tan}}
\index{aedl}{asin@{\tt asin}}
\index{aedl}{acos@{\tt acos}}
\index{aedl}{atan@{\tt atan}}
\index{aedl}{log@{\tt log}}
\index{aedl}{log10@{\tt log10}}
\index{aedl}{exp@{\tt exp}}
\index{aedl}{sqrt@{\tt sqrt}}
% 
These functions can be used everywhere, where
otherwise only a scalar number would appear. However, they can not be
applied to lists, such as {\tt sin(\{1,2,3\})} will not work.

@@

In \GLOBES\ 3.0 and higher, a number of new features can be used. The
features in this section are slighlty more advanced and can be jumped over at a first reading of
the manual. The
most important one are lists as variables in \AEDL . They start with {\tt \%}, such as
\begin{quote}
{\tt
\%effs = \{ 0.2, 0.4, 0.6, 0.8, 1.0, 1.0, 1.0 \}
}
\end{quote}
Functions can be threaded over lists, \ie, by applying the function to each element
and returning a list. Note that the original list will be destroyed by this
process. Therefore, it is necessary to create a copy of your list if you want to
use the original and the threading result. For that purpose, the {\tt copy}
funcion is provided:
\begin{quote}
{\tt
 listb = copy(lista) \\
 listb := lista // Alternative method 
}
\end{quote}
\index{aedl}{copy@{\tt copy}}
You will also need to use {\tt copy} when you assign a list to an experiment structure (see below).

Two helper functions {\tt bincenter()} 
\index{aedl}{bincenter@{\tt bincenter}} and {\tt simbincenter} \index{aedl}{simbincenter@{\tt simbincenter}} return lists with the central energies of the bins or sampling points, respectively. For example,
\begin{quote}
{\tt
 \%bc=bincenter() 
}
\end{quote}
A very useful new features is an interpolation function which can directly interpolate a number of
points and evaluate them at a different set of places. For example,
\begin{quote}
{\tt
 \%energ = \{ 4.0,20.0,50.0 \} \\
 \%effs = \{ 0.0,1.0,1.0 \} \\
 \%ires = interpolation(\%energ,\%effs,1,\%bc) 
}
\end{quote}
\index{aedl}{interpolation@{\tt interpolation}}
interpolates the points with x-values {\tt \%energ} and y-values {\tt \%effs} with the interpolation order one (linear interpolation, third parameter) and evaluates the interpolation result at the bin centers obtained above, \ie, it returns a list of the y-values at the places specified by the last parameter. The only allowed interpolation orders are ``1'' (linear) and ``2'' (cubic splines). This example creates
a neutrino factory energy threshold function linearly climbing from 0 to 1 between 4~GeV and 20~GeV. It can be directly used in a channel definition, \eg, 
\begin{quote}
{\tt
 channel(\#nu\_mu\_appearance)< \\
\hspace*{0.5cm}	@channel = \#mu\_plus: +: electron: muon: \#CC: \#MINOS\ \\
\hspace*{0.5cm}	@post\_smearing\_efficiencies = copy(\%ires) \\
>
}
\end{quote}

For debugging purposes of lists and numbers, one can \GLOBES\ now supports output directly
from \AEDL\ files:
\begin{quote}
{\tt 
 R = 1.15 \\
 echo(R) // Print without line feed \\
 line(2) // Two line feeds \\
 echon(ires) // Print with line feed 
}
\end{quote}
\index{aedl}{echo@{\tt echo}}
\index{aedl}{echon@{\tt echon}}
\index{aedl}{line@{\tt line}}

>>>>

Here, the general concept of the AEDL is described and illustrated by an example. In addition, a short introduction to the syntax of the AEDL is given.

@@

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method} (see, \eg, \Ref~\cite{Ohlsson:1999um})  to compute the neutrino oscillation probabilities and divide the matter density profile into layers of constant matter density. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained as the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even if only a small number of matter density steps is used. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

While it is comparatively simple to define a general neutrino source 
and to compute the oscillation physics, the general properties of a detector simulation are much more complicated. The basic assumption in building an abstract detector description is \emph{linearity}, \ie , that two neutrino events do not interfere with each other. Furthermore it is assumed that all information on the oscillation physics 
is given by the \emph{reconstructed} flavor and energy of a 
neutrino event. The term ``reconstructed'' implies that the well-defined energy of the incident neutrino, which can not be directly observed, translates via secondary particles and the detection properties into a distribution of possible energy values. This process is illustrated in \figu{distro} for the energy variable. 
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.6\textwidth]{mapping}
\end{center}
\mycaption{\label{fig:distro} A detector maps a true parameter value onto
a distribution of reconstructed parameter values. This is illustrated here for there energy.}
\end{figure}
% 
The same, in principle, applies to the nature of the neutrino flavor. However, in this case, only discrete values are applicable. Note that the reconstructed neutrino energy and the neutrino flavor are the only observables in \GLOBES .

@@

In order to implement an experiment definition including various
sources of systematical errors, we use several abstraction levels. 
The first level is the so-called ``channel'', \index{norm}{Channel}
 which is the link between 
the oscillation physics and the detection properties for a specfific oscillation pattern (\cf, \figu{channel}). A channel specifies the mapping of a specific neutrino flavor produced by the source onto a reconstructed neutrino flavor.
For example, a muon neutrino oscillates into an electron neutrino and subsequently interacts via quasi-elastic charged current scattering. The measured energy and direction of
the secondary electron in the detector then allows to reconstruct the neutrino energy. The connection from the source flux of the muon neutrino, via the  probability to appear as a electron neutrino, to its detection properties (such as cross sections and energy smearing) is encapsulated into the channel.

@@

The channels are the building blocks for the so-called ``rules''.
\index{norm}{Rule} In general, a rule consists of one or more ``signal'' and ``background'' oscillation channels, which are normalized with efficiencies
(\cf, \figu{rule}). The event numbers from these channels are added {\em before} the $\Delta \chi^2$-value is calculated.\footnote{Note that
in this manual, the $\chi^2$ and $\Delta \chi^2$ are equal, since
for simulated data $\Delta \chi^2 = 0$ at the best-fit point. Thus, we
are using $\chi^2$ and $\Delta \chi^2$ as equal quantities.} In addition, each rule implements {\em independent} systematics, such as signal and background normalization errors. Eventually, each rule gives a $\Delta \chi^2$-value, and the total $\Delta \chi^2$ of one experiment is obtained by adding the $\Delta \chi^2$'s of all rules (\cf, \figu{experiment}). 
 An example for a rule could look like this: We want to detect electron
  neutrino appearance (``signal''), where the overall efficiency for 
  quasi-elastics electron neutrino events is $0.4$. There is a fraction of 
  $0.01$ of all neutral current events which are mis-identified as 
  quasi-elastic electron neutrino events (``background''). The neutral 
  current fraction is only known within $10\%$ (``background uncertainty'') 
  and there is
an energy scale uncertainty of $100\,\mathrm{MeV}$ (``energy calibration error'').
All this systematics is independent of the other rules.  Thus, a rule connects the event rates to the calculation of a $\Delta \chi^2$ which properly includes systematical errors. The resulting $\Delta \chi^2$ is then the starting point for the oscillation physics analysis. Note again that
\begin{itemize}
\item
 Within each rule, the event numbers are added.
\item
 Within each rule, the systematics is treated independently from the other rules.
\item
 For each rule, the $\Delta \chi^2$ is computed; the $\Delta \chi^2$'s from all rules are added.
\end{itemize}

@@

Experiments in \GLOBES\ are defined by the Abstract Experiment Definition Language (AEDL). The experiment definition is written into a text file using the AEDL syntax. Currently, a number of pre-defined experiment definition files are provided with \GLOBES , which have to be modified manually in order to define new experiments.  The application software then uses this text file to initialize the experiment, where other secondary files might be read for source fluxes, cross sections \etc . In this section, we show the definition of a very simple neutrino factory in AEDL, where we do not go into details. In the next chapter, we will discuss each of the individual steps in detail.

@@

originating from stored $\mu^+$'s. This is achieved by setting the {\tt @builtin} variable to $1$. Next, we specify the muon energy to be $50\,\mathrm{GeV}$ by the {\tt @parent\_energy} variable. We assume 

@@

Then, we specify the file which contains the cross sections we want to 

@@

one element each, because we only use one density layer: We specify a
baseline length of $3000 \, \mathrm{km}$ with a constant matter density of
$3.5 \, \mathrm{g/cm^3}$. 

@@

above defined channel named {\tt \#appearance} with a constant overall

@@

The background error variable is defined in the same way as the signal error variable, \ie , we have a $20\%$ background uncertainty and a very small energy calibration error. In addition, the systematics treatment is specified

@@

The experiment defined here represents a first simplified version of a neutrino factory experiment. It still lacks the correct energy dependence of the efficiencies, the antineutrino disappearance channel, and the channels and rules for the symmetric operation with $\mu^-$ stored. However, it may serve as a simple, introductory example. In the next chapter, we will demonstrate that AEDL is much more powerful than illustrated here.

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages.  In addition, the minimum
\GLOBES\ version that the \AEDL\ file is supposed to run with has to be
defined by a {\tt \$version} statement, such as
{\tt \$version="3.0.0"}.
\index{aedl}{version@{\tt \$version}}
%
Comments can be used in the same way as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */ \\
// Another comment
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$.\footnote{The upper limit is 
only there for safety reasons, the memory is allocated dynamically.} If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environments in AEDL:
\begin{quote}
{\tt nuflux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a declaration without definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions, \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). In order to facilitate the the mapping from names in AEDL to indices
in C there are two functions \GLB{NameToValue} and \GLB{ValueToName} which
make this transition (see \Sec~\ref{sec:aedl_names}, 
page~\pageref{sec:aedl_names}).

@@

\index{aedl}{include@{\tt include}}
As another feature of AEDL one, can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "Exp1.glb"\\
        \#NEXT\#\\
        include   "Exp2.glb"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt Exp1.glb} and {\tt Exp2.glb}. Note that one has 
to use quotation marks for filenames in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively by two separate calls of 
\GLB{InitExperiment}.

Furthermore, one can define constants such as
\begin{quote}
{\tt
Pi = 3.14159
}
\end{quote}
These constants can not only be defined within one \AEDL\ file, but also
by the calling C program, which allows to use a simple but powerful variable
substitution mechanism as described in~\Sec~\ref{sec:aedlparams}.
\index{norm}{AEDL@\AEDL!external parameters}

In addition, some simple algebraic manipulations are possible, such as
\begin{quote}
{\tt
Pi+1\\
\verb+Pi^2+\\
sin(Pi/2)
}
\end{quote}
The following mathematical functions from {\tt <math.h>} are available: 
{\tt sin}, {\tt cos}, {\tt tan}, {\tt asin}, {\tt acos}, {\tt atan}, 
{\tt log}, {\tt log10}, {\tt exp}, {\tt sqrt}.
\index{aedl}{sin@{\tt sin}}%
\index{aedl}{cos@{\tt cos}}%
\index{aedl}{tan@{\tt tan}}%
\index{aedl}{asin@{\tt asin}}%
\index{aedl}{acos@{\tt acos}}%
\index{aedl}{atan@{\tt atan}}%
\index{aedl}{log@{\tt log}}%
\index{aedl}{log10@{\tt log10}}%
\index{aedl}{exp@{\tt exp}}%
\index{aedl}{sqrt@{\tt sqrt}}%
% 
These functions can be used everywhere, where
otherwise only a scalar number would appear. However, they can not be
applied to lists, \ie, expressions like {\tt sin(\{1,2,3\})} will not work.

@@

In \GLOBES\ 3.0 and higher, a number of new features can be used. The
features in this section are slighlty more advanced and can be skipped in a first reading of
the manual. The
most important one are lists as variables in \AEDL . They start with {\tt \%}, such as
\begin{quote}
{\tt
\%effs = \{ 0.2, 0.4, 0.6, 0.8, 1.0, 1.0, 1.0 \}
}
\end{quote}
Functions can be threaded over lists, \ie, they will be applied to each element
of list, and return a list. Note that the original list will be destroyed by this
process. Therefore, it is necessary to create a copy of your list if you want to
use the original and the threading result. For that purpose, the {\tt copy}
funcion is provided:
\begin{quote}
{\tt
 listb = copy(lista) \\
 listb := lista // Alternative method 
}
\end{quote}
\index{aedl}{copy@{\tt copy}}
You will also need to use {\tt copy} when you assign a list to an experiment structure (see below).

Two helper functions {\tt bincenter()} 
\index{aedl}{bincenter@{\tt bincenter}} and {\tt samplingbincenter} \index{aedl}{samplingbincenter@{\tt samplingbincenter}} return lists with the central energies of the bins or sampling points, respectively. For example,
\begin{quote}
{\tt
 \%bc=bincenter() 
}
\end{quote}
A very useful new features is an interpolation function which can directly interpolate a number of
points and evaluate them at a different set of places. For example,
\begin{quote}
{\tt
 \%energ = \{ 4.0,20.0,50.0 \} \\
 \%effs = \{ 0.0,1.0,1.0 \} \\
 \%ires = interpolation(\%energ,\%effs,1,\%bc) 
}
\end{quote}
\index{aedl}{interpolation@{\tt interpolation}}
interpolates the points with x-values {\tt \%energ} and y-values {\tt \%effs} with the interpolation order one (linear interpolation, third parameter) and evaluates the interpolation result at the bin centers obtained above, \ie, it returns a list of the y-values at the places specified by the last parameter. The only allowed interpolation orders are ``1'' (linear) and ``2'' (cubic splines). This example creates
a neutrino factory energy threshold function linearly climbing from 0 to 1 between 4~GeV and 20~GeV. It can be directly used in a channel definition, \eg, 
\begin{quote}
{\tt
 channel(\#nu\_mu\_appearance)< \\
\hspace*{0.5cm}	@channel = \#mu\_plus: +: electron: muon: \#CC: \#MINOS\ \\
\hspace*{0.5cm}	@post\_smearing\_efficiencies = copy(\%ires) \\
>
}
\end{quote}

To simplify debugging of lists and numbers, \GLOBES\ now supports output directly
from \AEDL\ files:
\begin{quote}
{\tt 
 R = 1.15 \\
 echo(R) // Print without line feed \\
 line(2) // Two line feeds \\
 echon(ires) // Print with line feed 
}
\end{quote}
\index{aedl}{echo@{\tt echo}}
\index{aedl}{echon@{\tt echon}}
\index{aedl}{line@{\tt line}}
